---
title: "Credit card fraud detection"
format: html
editor: visual
---

```{r echo=FALSE, warning=FALSE}
# Load necessary libraries
library(readr)
library(dplyr)
library(ggplot2)
```


```{r echo=FALSE, warning=FALSE}
# Load the dataset
data <- read_csv("data/creditcard.csv")

head(data)
```

## About the data

### Columns in the Credit Card Fraud Detection Dataset

| Column Name | Description |
|-------------|-------------|
| Time        | Seconds elapsed between this transaction and the first transaction in the dataset. |
| V1 to V28   | Anonymized features resulting from a PCA transformation. |
| Amount      | The transaction amount. |
| Class       | Target variable (0 for legitimate transactions, 1 for fraudulent transactions). |


## Exploring the data

```{r echo=FALSE, warning=FALSE}
# Summary statistics of the dataset
summary(data)

# Check for missing values
sum(is.na(data))

# Distribution of the target variable (Class)
table(data$Class)
```


```{r echo=FALSE, warning=FALSE}
# Plot the distribution of the Amount variable with improved aesthetics
ggplot(data, aes(x = Amount)) +
  geom_histogram(binwidth = 50, fill = "skyblue", color = "black", alpha = 0.7) +
  scale_x_continuous(limits = c(0, 2000), breaks = seq(0, 2000, by = 200)) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Distribution of Transaction Amounts", x = "Transaction Amount ($)", y = "Frequency") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```


```{r echo=FALSE, warning=FALSE}
# Plot the distribution of the Time variable
ggplot(data, aes(x = Time)) +
  geom_histogram(binwidth = 5000, fill = "green", color = "black") +
  labs(title = "Distribution of Transaction Times", x = "Time (seconds)", y = "Frequency")
```

